---
title: "recoverKBTDR"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit this file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- badges: start -->
`r badger::badge_license("GPL-3.0")`
`r badger::badge_last_commit("EmmaLiTsai/recoverKBTDR")`
`r badger::badge_github_actions("EmmaLiTsai/recoverKBTDR")` 
<!-- [![R-CMD-check](https://github.com/EmmaLiTsai/recoverKBTDR/workflows/R-CMD-check/badge.svg)](https://github.com/EmmaLiTsai/recoverKBTDR/actions) -->
<!-- badges: end -->

## Overview

This package contains a suite of functions for recovering dive records from 1970's film-based Kooyman-Billups Time Depth Records (KBTDRs). This TDR was among the first placed on a free-ranging marine organism, but the film-based format of the behavioral data makes long-term comparisons exceptionally challenging. This package returns a corrected, continuous, and digitized file of a KBTDR record complete with dates, times, and depth that can be easily read into dive analysis software. 

There are six main recovery steps that are achieved in this package: 

1. Scan centering and zero-offset correction 
2. Arc removal 
3. X-axis transformation to dates & times 
4. Interpolation between missing points 
5. Y-axis transformation to depth 
6. Spline smoothing 

## Installation

You can install the released version of recoverKBTDR from [CRAN](https://CRAN.R-project.org) with:

```{r,  load-data, message  = FALSE, results = 'hide'}
# load the development version: 
devtools::install_github("EmmaLiTsai/recoverKBTDR")
library(recoverKBTDR)
```
## Example of workflow

This is a basic example of the intended workflow for historic record recovery: 

First, read in data after scan image processing:
```{r reading-data}
# examples of data tidying from raw ImageJ csv files: 
filepath <- system.file("extdata", "WS_25_1981", package = "recoverKBTDR")
filepath_trace <- paste(filepath, "WS_25_1981_trace.csv", sep = "/")

tidy_trace <- tidy_raw_trace(filepath_trace)
head(tidy_trace)

# showing tidy data that is included within the package:
data(trace)
head(trace)
data(time_dots)
head(time_dots)
```

Center the scan using the timing dots, such that all timing dots will be centered along  y = -center_along_y. This helps remove drift in alignment due to record scanning:
```{r centering-data}
# centering along y = -0.9: 
trace <- center_scan(trace, time_dots, center_along_y = 0.9)

# visualizing the centered record: 
plot(tidy_trace[1000:11000,], xlab = "x position (cm)", ylab = "y position (cm)", type = "p", main = "Scan Centering") 
points(trace[1000:11000,], col = "#39b3b2")

# also extract the centered psi calibration curve: 
psi_calibration <- centered_psi_calibration(trace, psi_interval = c(100, 200, 400, 600, 800))

head(psi_calibration)
```

Some records present extreme drift and/or level shifts in surface values. This drift is common with modern TDRs and can be resolved using zero-offset correction methods modeled after the "diveMove" package. Code had to be modified from this package to handle the uncorrected trace data frame. These methods can be used with the zoc() function: 
```{r zoc}
trace_zoc <- zoc(trace, k_h  = 500, depth_bounds = c(-1, 1))

# visualizing this correction: 
plot(trace[1000:11000,],xlab = "x position (cm)", ylab = "y position (cm)", type = "p", main = "ZOC")
points(trace_zoc[1000:11000,], col = "#39b3b2")
```

Transform the x-axis into minutes using the timing dots, and remove the arc in the data by defining the height of the pivot point above y = 0 (center_y, in cm). 
```{r transform-x-axis}
# removing the data frame and also transform the x-axis to time in minutes from the origin. 
trace <- transform_x_vals(trace, 
                          time_dots, 
                          center_y = 11.18,
                          time_period_min = 12)

# visualizing:
plot(trace[1000:11000, 1:2], xlab = "x position (cm)", ylab = "y position (cm)", type = "p", main = "Arc Removal")
lines(trace[1000:11000, c(3,2)], col = "#39b3b2")

# to find the center_y value for arc removal, check out the helper function ?find_center_y() 
```

The seal often moved faster than the LED arm could document the dive during the descent and ascent phases. The function add_dates_times() uses the trace data frame to add a POSIXct date time object, and also interpolates between missing values to create a regular time series. 
```{r regular-time-series}
trace <- add_dates_times(trace,
                      start_time = "1981:01:16 15:10:00",
                      on_seal = "1981:01:16 17:58:00",
                      off_seal = "1981:01:23 15:30:00", 
                      tz = "Antarctica/McMurdo")
# now, the record is complete with POSIXct date times: 
head(trace)
```

Transform y-axis to depth either using the maximum depth, or psi calibration curve: 
```{r transform-to-depth}
# If psi calibration curve is present:
trace <- transform_y_vals(trace, psi_calibration = psi_calibration, max_psi = 900, max_position = 22.45)

# If we just know max depth:
trace_maxdep <- transform_y_vals(trace, max_dep = 319)

# visualizing this transformation: 
plot(trace[1000:18000, c(1,8)],xlab = "Date Time", ylab = "Depth (m)", type = "l", main = "Depth & Time Transformation")
```

Spline smoothing to reduce noise in the data by passing the spar value and depth threshold (in meters) to use when a dive is detected: 
```{r smoothing-data}
# smoothing the data frame with a rolling mean depth threshold of 5, and a spar value of 0.3 when a dive is detected: 
trace <- smooth_trace_dive(trace, spar_h = 0.3, depth_thresh = 5)

# seeing how smoothing preformed: 
plot(trace[1000:18000, c(1,8)], xlab = "Date Time", ylab = "Depth (m)", type = "l", main = "Smoothing Transformation", lwd = 3)
lines(trace[1000:18000, c(1,10)], col = "#39b3b2")

# to find the best spar value for spline smoothing, check out the helper function ?find_best_spar() 
```

Then, final data frame can be exported and read into dive analysis software. 

## Overall summary of workflow:
```{r workflow-summary}
library(recoverKBTDR)
## basic example code
data(trace) 
data(time_dots)

# centering the data
trace <- center_scan(trace, time_dots, center_along_y = 0.9)

# extracting the centered psi calibration curve, if needed: 
psi_calibration <- centered_psi_calibration(trace, psi_interval = c(100, 200, 400, 600, 800))

# arc removal and time assignment 
trace <- transform_x_vals(trace, time_dots, center_y = 11.18, time_period_min = 12)

# POSIXct date times and interpolating
trace <- add_dates_times(trace, start_time = "1981:01:16 15:10:00", on_seal = "1981:01:16 17:58:00", off_seal = "1981:01:23 15:30:00", tz = "Antarctica/McMurdo")

# add depths
trace <- transform_y_vals(trace, psi_calibration = psi_calibration, max_psi = 900, max_position = 22.45)

# spline smoothing 
trace <- smooth_trace_dive(trace, spar_h = 0.3, depth_thresh = 5)

# export! 

# fast recovery can also be used, where an argument csv file is used to quickly pass arguments to the functions above. For use, check out the ?fast_recovery(filepath)
```
